---
title: 'SG1022 Seminar 2: R BootCamp (2)'
author: "Christopher Gandrud"
date: "2016"
output:
  ioslides_presentation:
    css: https://maxcdn.bootstrapcdn.com/font-awesome/4.4.0/css/font-awesome.min.css
    logo: https://raw.githubusercontent.com/christophergandrud/Hertie_Collab_Data_Science/master/img/HertieCollaborativeDataLogo_v1.png
  beamer_presentation: default
  slidy_presentation: default
---

```{r include = FALSE}
library(dplyr)
```

## <i class="fa fa-arrow-circle-o-up"></i> Objectives

- Review R basics

- R example data sets

- Simple descriptive statistics,

- (Some) inferential statistics,

- Basic plots

- (A few) programming best practices

## <i class="fa fa-check"></i> Pop Quiz

## <i class="fa fa-table"></i> R's build-in data sets

List internal (example) data sets:

```{r eval=FALSE}
data()
```

Load **swiss** data set:

```{r}
data(swiss)
```

Find data description:

```{r, eval=FALSE}
?swiss
```

## <i class="fa fa-table"></i> R's build-in data sets

Find variable names:

```{r}
names(swiss)
```

See the first three rows and four columns

```{r}
head(swiss[1:3, 1:2])
```

## Note on Example data sets {.flexbox .vcenter}

Note, these dat sets are just for examples, you will **not use** them in your group projects.

## <i class="fa fa-area-chart"></i> Descriptive statistics: review

**Descriptive Statistics:** describe samples

Stats 101: describe sample **distributions** with appropriate measure of:

- **central tendancy**

- **variability**

## <i class="fa fa-area-chart"></i> Histograms

```{r}
hist(swiss$Examination)
```

## <i class="fa fa-area-chart"></i> Histograms: styling

```{r}
hist(swiss$Examination,
     main = 'Swiss Canton Draftee Examination Scores (1888)',
     xlab = '% receiving highest mark on army exam')
```

## Finding means

```{r}
mean(swiss$Examination)
```

If you have missing values (`NA`):

```{r}
mean(swiss$Examination, na.rm = TRUE)
```

## Other functions for central tendency

**Median**

```{r}
median(swiss$Examination)
```

**Mode**

`mode` is not an R function to find the statistical mode.

Instead use `summary` for factor nominal variables or make a bar chart.

## Simple bar chart for nominal

```{r, message=FALSE, warning=FALSE, cache=TRUE}
devtools::source_url('http://bit.ly/OTWEGS')
plot(MortalityGDP$region, xlab = 'Region')
```

## Using `summary`

You can also use the `summary` function to examine your data.

```{r}
summary(swiss$Examination)
```


## Variation

Variation is "perhaps the **most important quantity** in statistical analysis. The greater the variability in the data, the greater will be our **uncertainty** in the values of the parameters estimated . . . and the **lower our ability to distinguish between competing hypotheses**" (Crawley 2005, 33)


## Variation

**Range:**

```{r}
range(swiss$Examination)
```

**Quartiles:**

```{r}
summary(swiss$Examination)
```

## Variation

**Interquartile Range ($IQR = Q_{3} - Q_{1}$):**

```{r}
IQR(swiss$Examination)
```

## Variation

**Boxplots:**

```{r}
boxplot(swiss$Examination, main = '% of Draftees with Highest Mark')
```

## Variation: Standard Deviation

Use standard deviation **($s$)** to put variance in terms of the mean:

$$s = \sqrt{s^2}$$

```{r}
sd(swiss$Examination)
```


## Playing with distributions

Simulated normally distributed data with SD of 30 and mean 50

```{r}
Normal30 <- rnorm(1e+6, mean = 50, sd = 30)
```

```{r, echo=FALSE, message=FALSE}
# Helpful information from StackExchange discussion
# http://stackoverflow.com/questions/3494593/shading-a-kernel-density-plot-between-two-points?lq=1
# Load zoo package
library(zoo)
library(ggplot2)

# Convert to data frame
Normal30 <- data.frame(Normal30)

# Find density for +/- 1 SD
DensityLower <- density(Normal30$Normal30, from = 20, to = 50)
DensityUpper <- density(Normal30$Normal30, from = 50, to = 80)

# Extract density values (y values)
YLower <- DensityLower$y
YUpper <- DensityUpper$y

# Extract x values
XLower <- DensityLower$x
XUpper <- DensityUpper$x

# Merge vectors into data frames
Lower <- data.frame(YLower, XLower)
Upper <- data.frame(YUpper, XUpper)

# Add zero variable so that the shaded area goes down to 0
Lower$Zero <- 0
Upper$Zero <- 0

# Upper and Lower colours
LowerC <- "#80B2CD"
UpperC <- "#4F6F80"

# Create density plot
ggplot(Normal30, aes(x = Normal30)) +
        geom_density() +
        geom_ribbon(data = Lower, aes(x = XLower, ymax = YLower, ymin = Zero),
                    fill = LowerC, alpha = 0.4) +
        geom_ribbon(data = Upper, aes(x = XUpper, ymax = YUpper, ymin = Zero),
                    fill = UpperC, alpha = 0.4) +
        geom_vline(xintercept = 50, colour = "red", linetype = "longdash") +
        annotate("text", x = 25, y = 0.015, label = "-1 SD", colour = LowerC) +
        annotate("text", x = 75, y = 0.015, label = "+1 SD", colour = UpperC) +
        scale_y_continuous(limits = c(0, 0.02)) +
        scale_x_continuous(breaks = c(-90, -50, 0, 20, 50, 80, 100, 150, 185),
                           labels = c(-90, -50, 0, 20, 50, 80, 100, 150, 185)) +
        xlab("\n Simulated Data") + ylab("Density\n") +
        theme_bw()
```

## Transform skewed data

Highly skewed data can be transformed to have a normal distribution.


Helps correct two violations of key assumptions: (a) non-linearity and (b) heteroskedasticity.

```{r, fig.height=3}
hist(swiss$Education, main = '')
```

## Preview: Natural log transformed skewed data

```{r}
hist(log(swiss$Education), main = "Swiss Education")
```

## Transformations

The natural log transformation is only useful for data that **does not contain zeros**.

See <http://robjhyndman.com/hyndsight/transformations/> for suggestions on other
transformations such as Box-Cox and Inverse Hyperbolic Sine.

## Joint distributions (continuous variables)

```{r}
plot(log(swiss$Education), swiss$Examination)
```

## Preview: Summarise with correlation coefficients

```{r}
cor.test(log(swiss$Education), swiss$Examination)
```

## Preview: Summarise with loess

```{r, message=FALSE}
ggplot2::ggplot(swiss, aes(log(Education), Examination)) +
    geom_point() + geom_smooth() + theme_bw()
```

## A more systematic introduction to ggplot2

<br>
<br>
<br>

"gg" means "Grammar of Graphics".

"2" just means that it is the second one.

## ggplot2 syntax

Each plot is made of **layers**. Layers include the coordinate system (x-y), points,
labels, etc.

Each layer has **aesthetics** (`aes`) including the x & y, size, shape, and colour.

The **main layer types** are called **geometrics** (`geom`). These include
lines, points, density plots, bars, and text.

## ggplot2 examples setup

```{r, message=FALSE, warning=FALSE}
library(devtools)
library(ggplot2)

source_url("http://bit.ly/OTWEGS")

# Create data with no missing values of infant mortality
InfantNoMiss <- subset(MortalityGDP,
                           !is.na(InfantMortality))

# Create High/Low Income Variable
InfantNoMiss$DumMort[InfantNoMiss$InfantMortality
                     >= 15] <- "high"
InfantNoMiss$DumMort[InfantNoMiss$InfantMortality
                     < 15] <- "low"
```

## Simple example

```{r, warning=FALSE}
ggplot(data = MortalityGDP, aes(x = InfantMortality,
        y = GDPperCapita)) + geom_point()
```

## Simple example with BW theme

```{r warning=FALSE}
ggplot(data = MortalityGDP, aes(x = InfantMortality,
        y = GDPperCapita)) + geom_point() + theme_bw(base_size = 13)
```

## Colours

<br>
<br>

There are a number of ways to specify colours in ggplot2.

The simplest way is to let ggplot choose the colours for you.

---

```{r warning=FALSE}
ggplot(data = InfantNoMiss, aes(log(InfantMortality),
                                log(GDPperCapita))) +
      geom_point(aes(colour = income)) +
      theme_bw()
```

## Selecting colours

There are many ways to pick specific colors.

In this class we will mainly use **hexadecimal** colours.

This is probably the most commonly used system for choosing
colours on the web.

Every colour is given six digits.

A good website for getting hexadecimal colour schemes is:
<http://colorbrewer2.org/>.

---

```{r eval=FALSE}
# Create colour vector
Colours <- c("#1B9E77", "#D95F02", "#7570B3",
             "#E7298A", "#66A61E", "#E6AB02")

# Create graph
ggplot(data = InfantNoMiss,
                    aes(log(InfantMortality),
                        log(GDPperCapita))) +
        geom_point(aes(colour = income)) +
        scale_color_manual(values = Colours) +
        xlab("\nLog Infant Mortality") +
        ylab("Log GDP/Capita\n") +
        ggtitle("Log Transformed Data\n") +
        theme_bw()
```

---

```{r message=FALSE, echo=FALSE, warning=FALSE}
# Create colour vector
Colours <- c("#1B9E77", "#D95F02", "#7570B3",
             "#E7298A", "#66A61E", "#E6AB02")
# Create graph
ggplot(data = InfantNoMiss,
                    aes(log(InfantMortality),
                        log(GDPperCapita))) +
        geom_point(aes(colour = income)) +
        scale_color_manual(values = Colours) +
        xlab("\nLog Infant Mortality") +
        ylab("Log GDP/Capita\n") +
        ggtitle("Log Transformed Data\n") +
        theme_bw()
```

## ggplot2 is very flexible

```{r eval=FALSE}
# Create a violin Plot
ggplot(InfantNoMiss, aes(factor(DumMort),
                        log(GDPperCapita))) +
          geom_violin(fill = "#E7298A",
                      colour = "#E7298A",
                      alpha = I(0.5)) +
          geom_jitter(color = "#7570B3") +
          xlab("\n Infant Mortality") +
          ylab("Log GDP Per Capital\n") +
          theme_bw(base_size = 16)
```

---

```{r message=FALSE, echo=FALSE, warning=FALSE}
# Create a violin Plot
ggplot(InfantNoMiss, aes(factor(DumMort),
                        log(GDPperCapita))) +
          geom_violin(fill = "#E7298A",
                      colour = "#E7298A",
                      alpha = I(0.5)) +
          geom_jitter(color = "#7570B3") +
          xlab("\n Infant Mortality") +
          ylab("Log GDP Per Capital\n") +
          theme_bw(base_size = 16)
```

## Programming Hint (1)

Google is your friend!

If you have a question, someone else has probably asked it before.

## Programming Hint (2)

<i class="fa fa-exclamation"></i><i class="fa fa-exclamation">
</i><i class="fa fa-exclamation"></i> ***Always close!***

In R this means closing:

- `()`

- `[]`

- `{}`

- `''`

- `""`

## Programming Hint (3)

There are usually **many ways to achieve the same goal**, but . .

make your code as **simple as possible**.

- Easier to read.

- Easier to write (ultimately).

- Easier to find mistakes.

- Often computationally more efficient.

One way to do this is to **define things once**--e.g. use variables to contain
values and custom functions to contain multiple sequential function calls.

## Programming Hint (3)

**Bad**

```{r}
mean(rnorm(1000))

sd(rnorm(1000))
```

## Programming Hint (3)

**Good**

```{r}
rand_sample <- rnorm(1000)

mean(rand_sample)

sd(rand_sample)
```

## <i class="fa fa-arrow-circle-o-up"></i> Seminar: Use R!

- Load the `USArrests` data set:

```{r}
us_arrests <- USArrests
```

For more information use `?USArrests`

- Find appropriate summary statistics/summary plots for the variables in this data frame.

- Does there appear to be an association between a states' urban population and arrest rates?

# <i class="fa fa-arrow-up"></i> Advanced (if you're interested)

## Variation: Sum of Squares

**Sum of squares** (summing deviations from the mean):

$$\mathrm{Sum\:of\:Squares} = \sum(x - \bar{x})^2$$

- But sum of squares always gets bigger with a larger sample size.

    + Unless the new values exactly equal the mean.

## Variation: Degrees of Freedom

**Degrees of freedom** (number of values that are free to vary):

For the mean:

$$\mathrm{df} = n - 1$$

Why?

For a given mean and sample size, $n - 1$ values can vary, but the $n$th
value must always be the same. See Crawley (2005, 36-37).

## Variation: Variance

We can use degrees of freedom to create an "unbiased" measure of variability
that is not dependent on the sample size.

**Variance ($s^2$):**

$$
s^2 = \frac{\mathrm{Sum\:of\:Squares}}{\mathrm{Degrees\:of\:Freedom}} =
\frac{\sum(x - \bar{x})^2}{n - 1}
$$

But this is not in the same units as the mean, so it can be confusing to
interpret.

## Variation: Standard Deviation

Use standard deviation **($s$)** to put variance in terms of the mean:

$$s = \sqrt{s^2}$$

## Variation: Standard Error

The **standard error** of the mean:

If we think of the variation around a central tendency as a measure of the
**unreliability** of an estimate (mean) in a population,
then we want the measure to **decrease as the sample size goes up**.

$$
\mathrm{SE}_{\bar{x}} = \sqrt{\frac{s^{2}}{n}}
$$

Note: $\sqrt{}$ so that the dimensions of the measure of unreliability and the
parameter whose variability is being measured are the same.

Good overview of variance, degrees of freedom, and standard errors in Crawley
(2005, Ch. 4).

## Variation: Variance and Standard Deviation

**Variance:**

```{r}
var(swiss$Examination)
```

**Standard Deviation:**

```{r}
sd(swiss$Examination)
```

## Variation: Standard Error

**Standard Error:**

```{r}
sd_error <- function(x) {
    sd(x) / sqrt(length(x))
}

sd_error(swiss$Examination)
```

## <i class="fa fa-arrow-up"></i> Advance: Creating Functions

You can create a function to find the sample mean
($\bar{x} = \frac{\sum x}{n}$) of a vector.

```{r}
fun_mean <- function(x){
    sum(x) / length(x)
}

## Find the mean
fun_mean(x = swiss$Examination)
```

## <i class="fa fa-arrow-up"></i> Why create functions?

Functions:

- Simplify your code if you do repeated tasks.

- Lead to fewer mistakes.

- Are easier to understand.

- Save time over the long run--a general solution to problems in different
contexts.

## <i class="fa fa-arrow-up"></i> Advanced: Loops

You can 'loop' through the data set to find the mean for each column

```{r}
library(dplyr)

for (i in 1:length(names(swiss))) {
    swiss[, i] %>%
    mean() %>%
    round(digits = 1) %>%
    paste(names(swiss)[i], ., '\n') %>%  # the . directs the pipe
    cat()
}
```
