---
title: 'SG1022 Seminar 3: Using survey data'
author: "Christopher Gandrud"
date: '2016'
output:
  ioslides_presentation:
    css: https://maxcdn.bootstrapcdn.com/font-awesome/4.4.0/css/font-awesome.min.css
    logo: https://raw.githubusercontent.com/christophergandrud/Hertie_Collab_Data_Science/master/img/HertieCollaborativeDataLogo_v1.png
  pdf_document: default
---

## <i class="fa fa-arrow-circle-o-up"></i> Objectives

- Getting (local) data into R

- Working with factors (categorical data)

- Using R to analyse surveys (cross-tabs, $X^2$, grouped bar plots)

- Setting up Qualtrics

- Critique survey design

## <i class="fa fa-check"></i> Pop Quiz

- What is the `min-max` function and why would you use it when creating composite indicators?

- What is the `library` function and when do you use it?


## <i class="fa fa-folder-open-o"></i> File paths

Files are organised **hierarchically** into (upside down) trees.

```{r eval=FALSE}
Root
|_
    Parent
    |_
        Child1
        Child2
```

## Example file path

The file path `C:\group_project\data\data_set.csv` represents the tree:

```{r eval=FALSE}
C
|_
    group_project
    |_
        data
        |_
            data_set.csv

```


## <i class="fa fa-folder-open-o"></i> Root

**Root** directories are the **first level of a disk**.

They are the root out of which the file tree grows.

**Root naming conventions**:

**Linux/Mac**: `/`

- e.g. `/group_project` means that the `group_project` directory is a child
of the root directory.

**Windows**: the disk is partitioned, e.g. the `C` partition is denoted `C:\`.

- `C:\group_project` indicates that the `group_project` directory is a child
of the `C` partition.

## <i class="fa fa-folder-open-o"></i> Sub (child) directories

Sub (child) directories are denoted with a `/` in Linux/Mac and `\` in Windows, e.g.:

```{r engine='sh', eval=FALSE}
# Linux/Mac
/group_project/data

# Windows
C:\group_project\data
```

<i class="fa fa-exclamation"></i> **R tip**:

- In R for Windows you either use two backslashes `\\` (`\` is the R [escape
character](http://en.wikipedia.org/wiki/Escape_character))

- Or `/` in **relative paths** in R for Windows, it will know
what you mean.

## <i class="fa fa-folder-open-o"></i> Working directories

A **working directory** is the directory where the program looks for files/other
directories.

<i class="fa fa-exclamation"></i> **Always remember the working directory.**

- Otherwise you may open/save files that you do not want to open/save.

## <i class="fa fa-folder-open-o"></i> Working directories

**In R:**

```{r}
# Find working directory
getwd()

# List all files in the working directory
list.files()
```

```{r, eval=FALSE}
# Set group project as working directory
setwd('C:\\group_project')
```

## <i class="fa fa-folder-open-o"></i> Relative vs. Absolute file paths

Use **relative file paths** when possible.

- **Absolute file path**: the entire path on a particular system,
    + E.g. `C:\\group_project/data/data_set.csv`

- **Relative file path**: the path relative to the working directory.

    + E.g. if `C:\\group_project` is the working directory then the relative path for
    `data_set.csv` is `data/data_set.csv`.

Why?

- Your scripts will run easily on **other computers**. **Enhances reproducibility**.
Easier for your collaborators. Easier for you when you use another computer.

## <i class="fa fa-folder-open-o"></i> File & directory name conventions

<i class="fa fa-exclamation-triangle"></i> **Don't use spaces** in your file names.

They can create problems for programs that treat spaces as an indication that the
path has ended.

Alternatives:

- `CamelCase` (ex. `DataAnalysis.R`)

- `file_underscore` (ex. `data_analysis.R`)

## Loading data into R

R can load data from many different file formats (e.g. `.sav` (SPSS), `.xlsx` (Excel), `.dta` (Stata), `.csv`).

The [rio](https://github.com/leeper/rio) (R input/output) package makes it very easy to import many different types of data.

It has two key functions `import` and `export`.

**Remember:** You will need to `install.packages` rio and then load it into your workspace with `library`.

## Rio `import` for loading data into R

**First**, download the `ESS5_UKonly.sav` data set from Moodle (it's under Week 3). Create a folder called `sg1022_data` and place the data file into it.

**Second**, set your working directory to the folder where the data is located. For example, if your data is in `C:\\sg1022_data` then use:

```{r, eval=FALSE}
setwd('C:\\sg1022_data/')
```

## Rio `import` for loading data into R

Now load the data using the file's name (`ESS5_UKonly.sav`)

```{r message=FALSE, warning=FALSE}
library(rio)

# Load ESS 5 (UK) only data from SPSS format
ess5_uk <- import('ESS5_UKonly.sav')

# Show a selection of the data.
head(ess5_uk[1:3, 1:6])
```

# Note: Always look at the data you imported to see if it's what you think you imported/see what needs cleaning.

## Rio `export` for saving data

We can save the `ess5_uk` data into another format. For example, .csv "Comma Separated Values". Just add the csv file extension to the file name and `rio` does the rest.

```{r, eval=FALSE}
export(ess5_uk, 'ess5_uk.csv')
```

## Review: factors (categorical data)

```{r include=FALSE}
# Create fake data
#library(randomNames)

#persons <- randomNames(n = 1000)
#income <- sample(1:3, size = 1000, replace = T, prob = c(0.5, 0.4, 0.1))
#kanye_or_wiz <- sample(c('kanye', 'wiz'), size = 1000, replace = T, 
#                       prob = c(0.63, 1 - 0.63))
#ex_income <- data.frame(persons, income, kanye_or_wiz)
#export(ex_income, 'example_kw.csv')
```

In R, categorical data is coded using **factors**.

Let's load a simple data set into R for this example: `example_kw.csv`. 

This file is located on Moodle (Week 3) . Download it and place it into your `sg1022_data` directory. Now load it into a new object called `example`.

```{r message=FALSE}
example <- import('example_kw.csv')
```

## Review: factors (categorical data)

This data set has a numeric variable called `income`. We want to convert it to a factor with three category labels:

| Number Code | Label         |
| ----------- | ------------- |
| 1           | Low income    |
| 2           | Medium income |
| 3           | High income   |

## Factor labels

To convert this variable to a factor and add labels use the `factor` function.

```{r}
# Create a vector of level labels
income_labels <- c('Low income', 'Medium income', 'High income')

# Convert income to factor and apply labels
example$income <- factor(example$income, 
                               labels = income_labels)

# Create simple frequency table
summary(example$income)
```

## Converting from character strings to factors (1)

Sometimes you have data that is in character strings (R sees letters, but doesn't see any categories), but you want it to be a factor. For example:

```{r}
summary(example$kanye_or_wiz)
```

## Converting from character strings to factors (2)

Simply run it through `as.factor`. R will turn it into a factor and use the character strings as factor labels.

```{r}
example$kanye_or_wiz <- as.factor(example$kanye_or_wiz)

summary(example$kanye_or_wiz)
```

## Frequency table (categorical variables)

Remember, use `summary` to create a basic frequency table of a factor variable in R.

```{r, message=FALSE}
income_freq <- summary(example$income)
income_freq
```

We can convert these counts to proportions with `prop.table`:

```{r}
income_prop <- prop.table(income_freq)
income_prop
```


## Barplot (frequencies)

Show the frequencies more effectively with a barplot. To create a barplot of a single factor variable just use `plot`:

```{r}
plot(example$income, main = 'Income Level')
```

## Barplot (proportions)

You can also plot the proportions you created before. This time you need to explicitly use `barplot`.

```{r}
barplot(income_prop, main = 'Income Level', ylab = 'Proportion')
```

## Joint distributions (categorical variables)

Use `table` to create a simple contingency table:

```{r}
support <- table(example$kanye_or_wiz, example$income)
support
```

**Note:** the contingency table you create with the `table` function **is the basis for all** of the following cross-tabs, barplots, and $X^2$ tests.

## Cross-tabs with proportions

Use `prop.table` again to find the contingency table proportions:

```{r}
prop.table(support, margin = 1) # row proportions

prop.table(support, margin = 2) # column proportions
```

## Plotting (grouped bar chart)--Base R

```{r}
barplot(support, beside = TRUE, col = c('yellow', 'blue'),
  xlab = 'Income Level', ylab = 'Count',
  legend = rownames(support), args.legend = list(title = 'Support?'))
```

## Colours in R

For a list of R colour names see: <http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf>

## Plotting Percentages--Base R (1)

To plot proportions (easier for cross-group comparability), first create a table of the proportions:

```{r}
support_prop <- prop.table(support, margin = 2) # column proportion
support_prop
```

Then give these to `barplot` as before . . .

## Plotting Proportions--Base R (2)

```{r}
barplot(support_prop, beside = TRUE, col = c('yellow', 'blue'),
  xlab = 'Income Level', ylab = 'Proportion',
  legend = rownames(support), args.legend = list(title = 'Support?'))
```

## Plotting Percentages--Base R (1)

To plot percentages, simply convert the proportions to percents by multiplying them by 100:

```{r}
# Convert to percentages
support_perc <- support_prop * 100
support_perc
```

Then give these to `barplot` as before . . .

## Plotting Percentages--Base R (2)

```{r}
barplot(support_perc, beside = TRUE, col = c('yellow', 'blue'),
  xlab = 'Income Level', ylab = 'Percentage',
  legend = rownames(support), args.legend = list(title = 'Support?'))
```

## Joint distributions (categorical variables)

$\chi^2$ Test
```{r, warning=FALSE}
chisq.test(support)
```

Based on this test: is there a statistically significant association between income level and support for Kanye West and Wiz Khalifa?

## <i class="fa fa-arrow-circle-o-up"></i> Seminar: Work with example survey data

1. Set up a Qualtrics Survey using *Worksheet: introduction to Qualtrics* on Moodle under *Qualtrics Resources*.

2. With a partner complete the survey design worksheet.


## <i class="fa fa-arrow-up"></i> Extras: reorder factors

You may want to **change the order** of a factor's variable's levels so that makes more substantive sense. For example:

```{r}
library(MASS) # Contains example data set

# Relevel
survey$Smoke <- factor(survey$Smoke,
                       levels = c('Never', 'Occas', 'Regul', 'Heavy'))
survey$Exer <- factor(survey$Exer, levels = c('None', 'Some', 'Freq'))

# Create contingency table
smoking <- table(survey$Smoke, survey$Exer)
```
