---
title: 'SG1022 Seminar 3: Using survey data'
author: "Christopher Gandrud"
date: '2016'
output:
  ioslides_presentation:
    css: https://maxcdn.bootstrapcdn.com/font-awesome/4.4.0/css/font-awesome.min.css
    logo: https://raw.githubusercontent.com/christophergandrud/Hertie_Collab_Data_Science/master/img/HertieCollaborativeDataLogo_v1.png
---

## <i class="fa fa-arrow-circle-o-up"></i> Objectives

- Working directories

- Loading files with the `rio` package.

- Working with factors (categorical data)

- Using R to analyse surveys (contingency tables, )

## <i class="fa fa-check"></i> Pop Quiz

## Factors (categorical data)

In R, categorical data is coded using *factors*.

Imagine we have a data with three categories:

| Number Code | Label         |
| ----------- | ------------- |
| 1           | Low income    |
| 2           | Medium income |
| 3           | High income   |

## Factors

Imagine data is loaded into R just with the numbers:

```{r include=FALSE}
library(randomNames)
persons <- randomNames(n = 1000)
income <- sample(1:3, size = 1000, replace = T)
ex_income <- data.frame(persons, income)
```

```{r}
head(ex_income)
```

## Factor labels

To convert this variable to a factor and add labels use `factor`.

```{r}
income_labels <- c('Low income', 'Medium income', 'High income')

ex_income$income <- factor(ex_income$income, labels = income_labels)

summary(ex_income$income)
```

## Frequency table (categorical variables)

```{r, message=FALSE}
library(MASS)
smoking_freq <- table(survey$Smoke)
smoking_freq
```

## Joint distributions (categorical variables)

Contingency table
```{r}
smoking <- table(survey$Smoke, survey$Exer)
smoking

```

## Cross-tabs with proportions

```{r}
prop.table(smoking, margin = 1) # row proportions

prop.table(smoking, margin = 2) # column proportions
```

## Plotting (grouped bar chart)

```{r}
library(ggplot2)
ggplot(survey, aes(Smoke, fill = Exer)) +
    geom_bar(position = 'dodge')
```

## Joint distributions (categorical variables)

$\chi^2$ Test
```{r, warning=FALSE}
chisq.test(smoking)

```

## <i class="fa fa-arrow-circle-o-up"></i> Seminar: Work with example survey data
